Метод штрафов для решения задачи оптимального управления для квазилинейного параболического уравнения

### 2022-10-23 phd_4/article.tex
translation + cleanup

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[russian]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mhchem}
\usepackage{stmaryrd}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, filecolor=magenta, urlcolor=cyan,}
\urlstyle{same}
\usepackage{bbold}

\title{Penalty method to solve an optimal control problem for a qasilinear parabolic equation }

\author{(c) A. Yu. Chebotarev ${ }^{1,2}$, N. M. $\operatorname{Park}^{2,3}$, P. R. Mesenev, ${ }^{2}$ A. E. Kovtanyuk ${ }^{2,4}$}
\date{}

\begin{document}
\maketitle
$\mathrm{UDC} \quad 517.95$

\begin{abstract}
An optimal control problem for a quasilinear parabolic equation simulating the radiative and conductive heat transfer in a bounded three-dimensional domain under constraints on the solution in a given subdomain is considered. The solvability of the optomal control problem is proved. An algorithm for solving the problem, based on the penalty method, is proposed.
\end{abstract}

Key words: Non-linear PDE system, radiative heat transfer, optimal control, penalty method

DOI: \href{https://doi.org/10.47910/FEMJ2022XX}{https://doi.org/10.47910/FEMJ2022XX}

\section{Introduction}
The procedure of endovenous laser ablation (EVLA) is safe and sufficiently effective in the treatment of varicose veins. During EVLA, a laser optical fiber is inserted into the damaged vein. Then the laser radiation is transmitted through the fiber which at this time is pulled out of the vein. The end of the optical fiber is usually covered with a carbonized layer (optical fiber tip). The carbonized layer divides the laser energy into the fiber tip heating and radiation. The heat from the fiber tip is transmitted through the blood and surrounding tissue by the conductive heat transfer. The heat exchange is significantly increased by the flow of bubbles formed at the heated fiber tip. The radiation entering the blood and surrounding tissue is partially absorbed with the release of heat. As a result, the generated thermal energy causes significant heating of the vein which leads to its obliteration.

Mathematical modeling of radiation and thermal processes arising during EVLA is important to determine optimal parameters of radiation that provide a sufficiently high

${ }^{1}$ Institute for Applied Mathematics, Far Eastern Branch, Russian Academy of Sciences, Russia, 690041, Vladivostok, Radio st., 7.

${ }^{2}$ Far Eastern Federal University, Center for Research and Education in Mathematics, Russia, 690950, Vladivostok, Sukhanova str., 8.

${ }^{3}$ Amur State University, Russia, 675027, Blagoveshchensk, Amur region, Ignatyevskoye Shosse, 21.

${ }^{4}$ Technical University of Munich, Germany, 85748, München, Boltzmann str., 3. E-mail: \href{mailto:chebotarev.ayu@dvfu.ru}{chebotarev.ayu@dvfu.ru} (A. Yu. Chebotarev), \href{mailto:pak.nm@dvfu.ru}{pak.nm@dvfu.ru} (N. M. Park), \href{mailto:mesenev_pr@dvfu.ru}{mesenev\_pr@dvfu.ru} (P.R. Mesenev), \href{mailto:kovtanyuk.ae@dvfu.ru}{kovtanyuk.ae@dvfu.ru} (A. E. Kovtanyuk). temperature inside the vein for successful obliteration, on the other hand, the generated temperature field should be relatively safe for the live tissue surrounding the vein.

Mathematical model of EVLA takes into account the conductive heat transfer, as well as the radiation transfer and absorption with heat release. The flow of bubbles formed at the heated fiber tip makes a significant contribution to the temperature distribution in the model domain. In [1], based on the evaluation of experimental data, the heat transfer by the flow of bubbles is modeled using a piecewise constant thermal conductivity coefficient which depends on temperature as follows: when the temperature at some point reaches $95^{\circ} \mathrm{C}$, the coefficient of thermal conductivity increases 200 times.

Optimal control problems for the mathematical model of EVLA are studied in [2,3]. In [2] an optimal control problem of EVLA is posed, which consists in approximation of a given temperature profile at a certain point of the model domain. In [3], the similar as in [2] optimal control problem is studied. Here, the objective functional is taken such that its minimization allows one to reach the given temperature distribution in different parts of the model domain. This makes it possible to provide a sufficiently high temperature inside the vein for its successful obliteration and a safe temperature in the perivenous tissue. The unique solvability of the initial-boundary value problem is proved, on the basis of which the solvability of the optimal control problem is shown. An algorithm for finding a solution of the optimal control problem is proposed. Its efficiency is illustrated by a numerical example.

In the current work, an optimal control problem for the model of endovenous laser ablation in a bounded three-dimensional domain $\Omega$ with reflecting boundary $\Gamma=\partial \Omega$ is considered. The problem is to minimize the functional
$$
J(\theta)=\int_{G_{d}}\left(\left.\theta\right|_{t=T}-\theta_{d}\right)^{2} d x \rightarrow \inf
$$
on solutions of the initial-boundary value problem:
$$
\begin{gathered}
\sigma \partial \theta / \partial t-\operatorname{div}(k(\theta) \nabla \theta)-\beta \varphi=u_{1} \chi, \quad-\operatorname{div}(\alpha \nabla \varphi)+\beta \varphi=u_{2} \chi, \quad x \in \Omega, \quad 0<t<T, \\
k(\theta) \partial_{n} \theta+\left.\gamma\left(\theta-\theta_{b}\right)\right|_{\Gamma}=0, \quad \alpha \partial_{n} \varphi+\left.0.5 \varphi\right|_{\Gamma}=0,\left.\quad \theta\right|_{t=0}=\theta_{0} .
\end{gathered}
$$
In this case, the following restrictions are set:
$$
u_{1,2} \geq 0, \quad u_{1}+u_{2} \leq P,\left.\quad \theta\right|_{G_{b}} \leq \theta_{*} .
$$
Here, $\theta$ is the temperature, $\varphi$ the radiation intensity averaged over all directions, $\alpha$ the diffusion coefficient for optical radiation, $\mu_{a}$ the absorption coefficient, $k(\theta)$ the coefficient of thermal conductivity, $\sigma(x, t)$ the product of the specific heat capacity and the volume density, $u_{1}$ describes the power of the source spending on heating the fiber tip, $u_{2}$ is the power of the source spending on radiation, $\chi$ is equal to the characteristic function of the part of the medium in which the fiber tip is located divided by the volume of the fiber tip. The functions $\theta_{b}, \theta_{0}$ define the boundary and initial temperature distributions. We denote by $\partial_{n}$ the derivative in the direction of the outward normal $\mathbf{n}$ to the boundary $\Gamma$. It is required to provide the closeness of the temperature distribution to a desired temperature field $\theta_{d}$ at the final time $t=T$ in the $G_{d}$ subdomain, while the temperature in the subdomain $G_{b}$ does not exceed a constant critical value $\theta_{*}$.

\section{Formalization of the optimal control problem}
In what follows, we assume that $\Omega$ is a Lipschitz bounded domain, $\Gamma=\partial \Omega, Q=\Omega \times(0, T)$, $\Sigma=\Gamma \times(0, T)$. We denote by $L^{p}, 1 \leq p \leq \infty$, the Lebesgue space and by $H^{1}$ the Sobolev space $W_{2}^{1}$. The space $L^{p}(0, T ; X)$ (respectively, $C([0, T] ; X)$ ) consists of $p$-integrable on $(0, T)$ (respectively, continuous on $[0, T])$ functions with values in a Banach space $X$. Denote $H=L^{2}(\Omega), V=H^{1}(\Omega)$, and $V^{\prime}$ the dual of $V$. Then we identify $H$ with its dual space $H^{\prime}$ such that $V \subset H=H^{\prime} \subset V^{\prime}$, and denote by $\|\cdot\|$ the norm in $H$, and by $(h, v)$ the value of functional $h \in V^{\prime}$ on the element $v \in V$ coinciding with the inner product in $H$ if $h \in H$.

Let the following conditions hold:

@@

(iv) $0<\alpha_{0} \leq \alpha(x) \leq \alpha_{1}, \quad 0<\beta_{0} \leq \beta(x) \leq \beta_{1}, \quad x \in \Omega$

We define a nonlinear operator $A: V \rightarrow V^{\prime}$ and linear operator $B: V \rightarrow V^{\prime}$ using the following equality valid for any $\theta, v, \varphi, w \in V$ :
$$
(A(\theta), v)=(k(\theta) \nabla \theta, \nabla v)+\int_{\Gamma} \gamma \theta v d \Gamma=(\nabla h(\theta), \nabla v)+\int_{\Gamma} \gamma \theta v d \Gamma,
$$
where
$$
h(s)=\int_{0}^{s} k(r) d r ; \quad(B \varphi, w)=(\alpha \nabla \varphi, \nabla w)+(\beta \varphi, w)+\frac{1}{2} \int_{\Gamma} \varphi w d \Gamma
$$
Further, by the following bilinear form, we define the inner product in $V$ :
$$
(u, v)_{V}=(\nabla u, \nabla v)+\int_{\Gamma} u v d \Gamma .
$$
The corresponding norm is equivalent to the standard norm of the space $V$.

Definition 1. Let $u_{1,2} \in L^{2}(0, T)$. The pair $\theta, \varphi \in L^{2}(0, T ; V)$ is a weak solution of the problem (1), (2) if $\sigma \theta^{\prime} \in L^{2}\left(0, T ; V^{\prime}\right)$ and
$$
\sigma \theta^{\prime}+A(\theta)-\beta \varphi=g+u_{1} \chi, \quad \theta(0)=\theta_{0}, \quad B \varphi=u_{2} \chi,
$$
where
$$
\theta^{\prime}=d \theta / d t, \quad g \in L^{\infty}\left(0, T ; V^{\prime}\right), \quad(g, v)=\int_{\Gamma} \gamma \theta_{b} v d \Gamma
$$
Remark 1. Since $(\sigma \theta)^{\prime}=\sigma \theta^{\prime}+\theta \partial \sigma / \partial t \in L^{2}\left(0, T ; V^{\prime}\right)$ and $\sigma \theta \in L^{2}(0, T ; V)$, then $\sigma \theta \in$ $C([0, T] ; H)$, and therefore the initial condition makes sense.

It follows from the Lax-Milgram lemma that for any function $g \in H$ there is a unique solution of equation $B \varphi=g$. Moreover, the inverse operator $B^{-1}: H \rightarrow V$ is continuous. Therefore, we can exclude the radiation intensity $\varphi=u_{2} B^{-1} \chi$ and formulate the optimal control problem as follows. Problem (CP)
$$
\begin{gathered}
J(\theta)=\int_{G_{d}}\left(\left.\theta\right|_{t=T}-\theta_{d}\right)^{2} d x \rightarrow \inf , \quad \sigma \theta^{\prime}+A(\theta)=g+u, \quad \theta(0)=\theta_{0}, \\
\left.\theta\right|_{G_{b}} \leq \theta_{*}, \quad u \in U_{a d} .
\end{gathered}
$$
Here,
$$
U_{a d}=\left\{u=u_{1} \chi+u_{2} \beta B^{-1} \chi: u_{1,2} \in L^{2}(0, T), u_{1,2} \geq 0, u_{1}+u_{2} \leq P\right\}
$$

\section{Preliminary results}
In the article [5], the following result is obtained.

Lemma 1. Let conditions (i) - (iv) hold and $u \in L^{2}\left(0, T ; V^{\prime}\right)$. Then there is a solution to the problem
$$
\sigma \theta^{\prime}+A(\theta)=g+u, \quad \theta(0)=\theta_{0},
$$
such that $\theta \in L^{\infty}(0, T ; H)$ and the following estimate is valid:
$$
\|\theta(t)\|^{2}+\|\theta\|_{L^{2}\left(0, T ; V^{\prime}\right)}^{2} \leq C\left(\left\|\theta_{0}\right\|^{2}+\|g+u\|_{L^{2}\left(0, T ; V^{\prime}\right)}^{2}\right),
$$
where $C>0$ does not depend on $\theta_{0}, g$, and $u$.

Lemma 2. Let conditions (i) - (iv) hold, $u=0, \theta_{0} \leq \theta_{*}$ a.e. in $\Omega, \theta_{b} \leq \theta_{*}$ a.e. in $\Sigma$, and $\theta$ be a solution to the problem (4). Then $\theta \leq \theta_{*}$ a.e. in $\Omega \times(0, T)$.

Proof. Multiplying in the sense of inner product in $H$ the first equation in (4) by $v=\max \left\{\theta-\theta_{*}, 0\right\} \in L^{2}(0, T ; V)$, we get
$$
\left(\sigma v^{\prime}, v\right)+(k(\theta) \nabla v, \nabla v)+\int_{\Gamma} \gamma \theta v d \Gamma=0 .
$$
Discarding the nonnegative second and third terms, we arrive at the estimate
$$
\frac{d}{d t}(\sigma v, v) \leq\left(\sigma_{t} v, v\right) \leq \sigma_{2}\|v\|^{2} .
$$
Integrating the last inequality with respect to time and taking into account that $\left.v\right|_{t=0}=0$, we obtain
$$
\sigma_{0}\|v(t)\|^{2} \leq(\sigma v(t), v(t)) \leq \sigma_{2} \int_{0}^{t}\|v(\tau)\|^{2} d \tau
$$
Based on the Gronwall lemma, we conclude that $v=0$ and therefore $\theta \leq \theta_{*}$ a.e. in $\Omega \times(0, T)$

Lemmas 1 and 2 imply a non-empty set of admissible pairs of the Problem (CP) and the boundedness of a minimizing sequence of admissible pairs $\left\{\theta_{m}, u_{m}\right\} \in L^{2}(0, T ; V) \times$ $U_{a d}$ such that $J\left(\theta_{m}\right) \rightarrow j=\inf J$, where
$$
\sigma \theta_{m}^{\prime}+A\left(\theta_{m}\right)=g+u_{m}, \quad \theta_{m}(0)=\theta_{0},\left.\quad \theta_{m}\right|_{G_{b}} \leq \theta_{*} .
$$
Similarly [4], passing to the limit in system (5), it is possible to establish the solvability of the Problem (CP).

Theorem 1. Let conditions (i) $-(i v)$ hold, $\theta_{0} \leq \theta_{*}$ a.e. in $\Omega, \theta_{b} \leq \theta_{*}$ a.e. in $\Sigma$. Then a solution of the Problem $(C P)$ exists.

\section{Penalty problem}
Let us consider the following optimal control problem with the parameter $\varepsilon>0$ whose solutions approximate the solution of the Problem (CP) as $\varepsilon \rightarrow+0$.

Problem $\left(\mathrm{CP}_{\varepsilon}\right)$
$$
\begin{gathered}
J_{\varepsilon}(\theta)=\int_{G_{d}}\left(\left.\theta\right|_{t=T}-\theta_{d}\right)^{2} d x+\frac{1}{\varepsilon} \int_{0}^{T} \int_{G_{b}} F(\theta) d x d t \rightarrow \inf \\
\sigma \theta^{\prime}+A(\theta)=g+u, \quad \theta(0)=\theta_{0}, \quad u \in U_{a d}
\end{gathered}
$$
Here,
$$
F(\theta)= \begin{cases}0, & \text { if } \theta \leq \theta_{*}, \\ \left(\theta-\theta_{*}\right)^{2}, & \text { if } \theta>\theta_{*}\end{cases}
$$
The estimates presented in Lemma 1 make it possible, similarly as in the proof of Theorem 1 , to prove the solvability of the problem with the penalty.

Theorem 2. Let conditions (i)-(iv) hold. Then a solution of the problem $\left(C P_{\varepsilon}\right)$ exists.

Consider the approximation properties of solutions to the problem with the penalty. Let $\left\{\theta_{\varepsilon}, u_{\varepsilon}\right\}$ be solutions to the Problem $\left(\mathrm{CP}_{\varepsilon}\right)$ and $\{\theta, u\}$ be a solution to the Problem (CP). Then,
$$
\sigma \theta_{\varepsilon}^{\prime}+A\left(\theta_{\varepsilon}\right)=g+u_{\varepsilon}, \quad \theta_{\varepsilon}(0)=\theta_{0} .
$$
Since $\left.\theta\right|_{G_{b}} \leq \theta_{*}$, the following inequalities are true:
$$
\int_{G_{d}}\left(\left.\theta_{\varepsilon}\right|_{t=T}-\theta_{d}\right)^{2} d x \leq J(\theta), \quad \int_{0}^{T} \int_{G_{b}} F\left(\theta_{\varepsilon}\right) d x d t \leq \varepsilon J(\theta) .
$$
From the estimates obtained, using if necessary subsequences as $\varepsilon \rightarrow+0$, similarly as in the proof of Theorem 1 , we can prove the existence of functions $\widehat{u} \in U_{a d}, \widehat{\theta} \in L^{2}(0, T ; V)$ such that

>>>>

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[russian]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mhchem}
\usepackage{stmaryrd}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, filecolor=magenta, urlcolor=cyan,}
\urlstyle{same}
\usepackage{bbold}
\usepackage{amsfonts}
\usepackage{amsmath}

\title{Метод штрафов для решения задачи оптимального управления для квазилинейного параболического уравнения}

\author{А.Ю. Чеботарёв, Н.М. Пак, П.Р. Месенёв, А.Е. Ковтанюк}
\date{}

\begin{document}
    \maketitle
    \begin{abstract}
        Рассматривается задача оптимального управления для квазилинейного параболического уравнения,
        моделирующего радиационный и кондуктивный теплообмен в ограниченной трехмерной области при
        ограничениях на решение в заданной области.
        Доказана разрешимость задачи оптимального управления.
        Предложен алгоритм решения задачи, основанный на методе штрафов.
    \end{abstract}

    Ключевые слова: Нелинейная PDE-система, радиационный теплообмен, оптимальное управление, метод штрафов.

    \section{Введение}
    Процедура эндовенозной лазерной абляции (EVLA) безопасна и достаточно эффективна при
    лечении варикозного расширения вен.
    Во время ЭВЛА в поврежденную вену вводится лазерное оптическое волокно.
    Затем лазерное излучение передается по волокну, которое в это время вытягивается из вены.
    Конец оптического волокна обычно покрыт карбонизированным слоем (наконечник оптического волокна).
    Карбонизированный слой разделяет лазерную энергию на нагрев кончика волокна и излучение.
    Тепло от наконечника волокна передается через кровь и окружающие ткани за счет кондуктивной теплопередачи.
    Теплообмен значительно увеличивается за счет потока пузырьков, образующихся на нагретом кончике волокна.
    Излучение, попадающее в кровь и окружающие ткани, частично поглощается с выделением тепла.
    В результате генерируемая тепловая энергия вызывает значительный нагрев вены, что приводит к ее облитерации.

    Математическая модель EVLA учитывает кондуктивную теплопередачу, а также передачу излучения
    и поглощение с выделением тепла.
    Поток пузырьков, образующихся на нагретом наконечнике волокна,
    вносит значительный вклад в распределение температуры в модельной области.
    В [1], основываясь на оценке экспериментальных данных, теплопередача потоком пузырьков моделируется
    с использованием кусочно-постоянного коэффициента теплопроводности, который зависит от температуры
    следующим образом: когда температура в некоторой точке достигает $95 ^ {\ circ} \mathrm {C}$,
    коэффициент теплопроводность увеличивается в 200 раз.

    Задачи оптимального управления для математической модели EVLA изучены в [2,3].
    В [2] ставится задача оптимального управления EVLA, которая заключается в
    аппроксимации заданного температурного профиля в определенной точке модельной области.
    В [3] изучается аналогичная [2] задача оптимального управления.
    Здесь целевой функционал берется таким образом, что его минимизация позволяет достичь
    заданного распределения температуры в разных частях модельной области.
    Это позволяет обеспечить достаточно высокую температуру внутри вены для ее успешной
    облитерации и безопасную температуру в перивенозной ткани.
    Доказана однозначная разрешимость начально-краевой задачи,
    на основе которой показана разрешимость задачи оптимального управления.
    Предложен алгоритм нахождения решения задачи оптимального управления.
    Его эффективность проиллюстрирована численным примером.

    В настоящей работе рассматривается задача оптимального управления для модели
    эндовенозной лазерной абляции в ограниченной трехмерной области
    $\Omega$ с отражающей границей $\Gamma=\partial\Omega$.
    Проблема состоит в том, чтобы свести к минимуму функционал
    \[
        J(\theta)=\int_{G_{d}}\left(\left.\theta\right|_{t=T}-\theta_{d}\right)^{2} d x \rightarrow \inf
    \]
    на решениях начально-краевой задачи
    \[
        \begin{gathered}
            \sigma \partial \theta / \partial t-\operatorname{div}(k(\theta) \nabla \theta)-\beta \varphi=u_{1} \chi, \quad-\operatorname{div}(\alpha \nabla \varphi)+\beta \varphi=u_{2} \chi, \quad x \in \Omega, \quad 0<t<T, \\
            k(\theta) \partial_{n} \theta+\left.\gamma\left(\theta-\theta_{b}\right)\right|_{\Gamma}=0, \quad \alpha \partial_{n} \varphi+\left.0.5 \varphi\right|_{\Gamma}=0,\left.\quad \theta\right|_{t=0}=\theta_{0} .
        \end{gathered}
    \]
    В этом случае устанавливаются следующие ограничения:
    \[
        u_{1,2} \geq 0, \quad u_{1}+u_{2} \leq P,\left.\quad \theta\right|_{G_{b}} \leq \theta_{*} .
    \]
    Здесь $\theta$ - температура, $\varphi$ - интенсивность излучения, усредненная по всем направлениям,
    $\alpha$ - коэффициент диффузии оптического излучения, $\mu_{a}$ - коэффициент поглощения,
    $k(\theta)$ - коэффициент теплопроводности, $\sigma(x, t)$ произведение удельной теплоемкости
    и объемной плотности, $u_{1}$ описывает мощность источника, затрачиваемую
    на нагрев наконечника волокна, $u_{2}$ - мощность источника, затрачиваемая на излучение,
    $\chi$ равно характеристической функции части среды, в которой расположен наконечник волокна,
    деленной на объем наконечника волокна.
    Функции $\theta_{b}, \theta_{0}$ определяют граничное и начальное распределения температуры.
    Мы обозначим через $\partial_{n}$ производную в направлении внешней нормали $\mathbf{n}$ к границе $\Gamma$.
    Требуется обеспечить близость распределения температуры к желаемому температурному полю $\theta_{d}$
    в конечный момент времени $t=T$ в поддомене $G_{d}$, при этом температура в поддомене $G_{b}$
    не превышает постоянного критического значения значение $\theta_{*}$.

    \section{Формализация задачи оптимального управления}

    В дальнейшем мы предполагаем, что $\Omega $ является ограниченной областью Липшица,
    $\Gamma =\partial \Omega, Q =\Omega \times(0, T)$, $\Sigma=\Gamma \times(0, T)$.
    Обозначим через $L ^ {p}, 1 \leq p \leq \infty$ пространство Лебега и через $H^{1}$
    пространство Соболева $W_{2}^{1}$.
    Пространство $L ^ {p}(0, T ; X)$ (соответственно, $C([0, T] ; X)$ ) состоит из
    $p$-интегрируемых по $(0, T)$ (соответственно, непрерывных по $[0, T])$ функции со значениями
    в банаховом пространстве $X$.
    Обозначим $H=L ^{2}(\Omega), V=H^{1} (\Omega)$ и $V ^ {\prime}$ двойственное значение $V$.
    Затем мы отождествляем $H $ с его двойным пространством $H ^ {\prime}$ таким,
    что $V \subset H = H ^ {\prime} \subset V ^ {\prime}$ и обозначаем
    через $ \|\cdot \|$ норму в $H$, и на $(h, v)$ значение функционала $h \in V ^ {\prime}$
    на элементе $v \ в V$, совпадающее с внутренним произведением в $H$, если $h \in H$.

    Пусть выполняются следующие условия:

@@

    (iv) $0<\alpha_{0} \leq \alpha(x) \leq \alpha_{1}, \quad 0<\beta_{0} \leq \beta(x) \leq \beta_{1}, \quad x \in \Omega$
    Мы определяем нелинейный оператор $A:V\rightarrow V^{\prime}$
    и линейный оператор $B:V\rightarrow V ^ {\prime}$,
    используя следующее равенство, действительное для любого $\theta, v, \varphi, w \in V$ :
    \[
        (A(\theta), v)=(k(\theta) \nabla \theta, \nabla v)+\int_{\Gamma} \gamma \theta v d
        \Gamma=(\nabla h(\theta), \nabla v)+\int_{\Gamma} \gamma \theta v d \Gamma,
    \]

    где%в пизде
    \[
        h(s)=\int_{0}^{s} k(r) d r ; \quad(B \varphi, w)=
        (\alpha \nabla \varphi, \nabla w)+(\beta \varphi, w)+\frac{1}{2} \int_{\Gamma} \varphi w d \Gamma
    \]
    Далее, с помощью следующей билинейной формы, мы определяем внутреннее произведение в$V$ :
    \[
        (u, v)_{V}=(\nabla u, \nabla v)+\int_{\Gamma} u v d \Gamma.
    \]
    Соответствующая норма эквивалентна стандартной норме пространства $V$.

    \textbf{Определение 1.} Пусть $u_{1,2} \in L^{2}(0, T)$.
    Пара $\theta, \varphi \in L^{2}(0, T ; V)$ слабое решение задачи (1), (2) если
    $\sigma \theta^{\prime} \in L^{2}\left(0, T ; V^{\prime}\right)$ и
    \[
        \sigma \theta^{\prime}+A(\theta)-\beta \varphi=g+u_{1} \chi, \quad \theta(0)=\theta_{0}, \quad B \varphi=u_{2} \chi,
    \]
    где
    \[
        \theta^{\prime}=d \theta / d t, \quad g \in L^{\infty}\left(0, T ; V^{\prime}\right),
        \quad(g, v)=\int_{\Gamma} \gamma \theta_{b} v d \Gamma
    \]
    \textit{Замечание 1.} Так как $(\sigma \theta)^{\prime}=\sigma \theta^{\prime}+\theta
    \partial \sigma / \partial t \in L^{2}\left(0, T ; V^{\prime}\right)$ and $\sigma
    \theta \in L^{2}(0, T ; V)$, then $\sigma \theta \in$ $C([0, T] ; H)$, и поэтому начальные условия имеют физические основания.

    Из леммы Лакса-Милграма следует, что для любой функции $g\ в H$ существует единственное
    решение уравнения $B\varphi= g$.
    Более того, обратный оператор $B ^{-1}: H \rightarrow V$ является непрерывным.
    Следовательно, мы можем исключить интенсивность излучения $\varphi=u_{2} B ^ {-1} \chi$
    и сформулировать задачу оптимального управления следующим образом.
    Проблема (CP)
    \[
        \begin{gathered}
            J(\theta)=\int_{G_{d}}\left(\left.\theta\right|_{t=T}-\theta_{d}\right)^{2} d x \rightarrow \inf ,
            \quad \sigma \theta^{\prime}+A(\theta)=g+u, \quad \theta(0)=\theta_{0}, \\
            \left.\theta\right|_{G_{b}} \leq \theta_{*}, \quad u \in U_{a d}.
        \end{gathered}
    \]
    Здесь
    \[
        U_{a d}=\left\{u=u_{1} \chi+u_{2} \beta B^{-1} \chi: u_{1,2} \in L^{2}(0, T), u_{1,2}
        \geq 0, u_{1}+u_{2} \leq P\right\}
    \]

    \section{Предварительные результаты}
    В статье [5] получен следующий результат.

    \textit{Лемма 1.}
    Пусть условия (i) - (iv) выполняются и $u \in L^{2}\left(0, T ; V^{\prime}\right)$.
    Тогда есть решение проблемы
    \[
        \sigma \theta^{\prime}+A(\theta)=g+u, \quad \theta(0)=\theta_{0},
    \]
    такое что $\theta \in L^{\infty}(0, T ; H)$, а также верна следующая оценка:
    \[
        \|\theta(t)\|^{2}+\|\theta\|_{L^{2}\left(0, T ; V^{\prime}\right)}^{2} \leq C\left(\left\|\theta_{0}\right\|^{2}+\|g+u\|_{L^{2}\left(0, T ; V^{\prime}\right)}^{2}\right),
    \]

    где $C>0$ не зависит от $\theta_{0}, g$, и $u$.

    \textit{Lemma 2.}
    Пусть условия (i) - (iv) выполняются, $u=0, \theta_{0} \leq \theta_{*}$ другими словами,
    в $\Omega, \theta_{b} \leq \theta_{*}$ то есть $\Sigma$, и $\theta$ будут решением задачи (4).
    Тогда $\theta \leq \theta_{*}$ в $\Omega \times(0, T)$.

    \textit{Доказательство.}
    Умножая в смысле внутреннего произведения в $H$ первое уравнение в (4) на
    $v=\max \left\{\theta-\theta_{*}, 0\right\}\in L^{2}(0, T; V)$, мы получаем
    \[
        \left(\sigma v^{\prime}, v\right)+(k(\theta) \nabla v, \nabla v)+\int_{\Gamma} \gamma \theta v d \Gamma=0 .
    \]

    Отбрасывая неотрицательные второе и третье слагаемые, мы приходим к оценке
    \[
        \frac{d}{d t}(\sigma v, v) \leq\left(\sigma_{t} v, v\right) \leq \sigma_{2}\|v\|^{2} .
    \]

    Интегрируя последнее неравенство по времени и принимая во внимание, что $\left.v\right|_{t=0}=0$, мы получаем
    \[
        \sigma_{0}\|v(t)\|^{2} \leq(\sigma v(t), v(t)) \leq \sigma_{2} \int_{0}^{t}\|v(\tau)\|^{2} d \tau
    \]
    Основываясь на лемме Гронуолла, мы приходим к выводу, что $v=0$ и, следовательно,
    $\theta \leq \theta_{*}$ в $\Omega \times(0, T)$

    Леммы 1 и 2 подразумевают непустое множество допустимых пар задачи (CP)
    и ограниченность минимизирующей последовательности допустимых пар
    $\left\{\theta_{m}, u_{m}\right\} \in L^{2}(0, T ; V) \times$
    $U_{a d}$ так, что $J\left(\theta_{m}\right) \rightarrow j=\inf J$, где
    \[
        \sigma \theta_{m}^{\prime}+A\left(\theta_{m}\right)=g+u_{m},
        \quad \theta_{m}(0)=\theta_{0},\left.\quad \theta_{m}\right|_{G_{b}} \leq \theta_{*}.
    \]
    Аналогично [4], переходя к пределу в системе (5), можно установить разрешимость задачи (CP).

    \textbf{Theorem 1.}
    Пусть условия (i)-(iv) выполняются,
    $\theta_{0} \leq \theta_{*}$ a.e. в $\Omega, \theta_{b} \leq \theta_{*}$ a.e. в $\Sigma$.
    Тогда решение проблемы (CP) существует.

    \section{Метод штрафов}
    Рассмотрим следующую задачу оптимального управления с параметром $\varepsilon>0$,
    решения которой аппроксимируют решение задачи (CP) как $\varepsilon \rightarrow+0$.
    Problem $\left(\mathrm{CP}_{\varepsilon}\right)$
    \[
        \begin{gathered}
            J_{\varepsilon}(\theta)=\int_{G_{d}}\left(\left.\theta\right|_{t=T}-\theta_{d}\right)^{2} d x+\frac{1}{\varepsilon} \int_{0}^{T} \int_{G_{b}} F(\theta) d x d t \rightarrow \inf \\
            \sigma \theta^{\prime}+A(\theta)=g+u, \quad \theta(0)=\theta_{0}, \quad u \in U_{a d}
        \end{gathered}
    \]
    Здесь,
    \[
        F(\theta)=
        \begin{cases}
            0, & \text { if } \theta \leq \theta_{*}, \\
            \left(\theta-\theta_{*}\right)^{2}, & \text { if } \theta>\theta_{*}
        \end{cases}
    \]
    Оценки, представленные в лемме 1, позволяют, аналогично доказательству теоремы 1, доказать разрешимость задачи со штрафом.
    \textbf{Теорема 2.} Пусть выполняются условия (i)-(iv).
    Тогда существует решение проблемы $\left(C P_{\varepsilon}\right)$.

    Рассмотрим аппроксимативные свойства решений задачи со штрафом.
    Пусть $\left\{\theta_{\varepsilon}, u_{\varepsilon}\right\}$ будет решением проблемы
    $\left(\mathrm{CP}_{\varepsilon}\right)$ и $\{\theta, u\}$ будет решением проблемы(CP).
    Тогда,
    \[
        \sigma \theta_{\varepsilon}^{\prime}+A\left(\theta_{\varepsilon}\right)=g+u_{\varepsilon},
        \quad \theta_{\varepsilon}(0)=\theta_{0}.
    \]

    так как $\left.\theta\right|_{G_{b}} \leq \theta_{*}$, верны следующие неравенства:
    \[
        \int_{G_{d}}\left(\left.\theta_{\varepsilon}\right|_{t=T}-\theta_{d}\right)^{2} d x \leq J(\theta),
        \quad \int_{0}^{T} \int_{G_{b}} F\left(\theta_{\varepsilon}\right) d x d t \leq \varepsilon J(\theta).
    \]

    Из полученных оценок, используя при необходимости подпоследовательности в качестве
    $\varepsilon \rightarrow+0$, аналогично, как и в доказательстве теоремы 1, мы можем доказать существование функций
    $\widehat{u} \in U_{a d}, \widehat{\theta} \in L^{2}(0, T ; V)$ таких, что

    $u_{\varepsilon} \rightarrow \widehat{u}$ слабо в
    $L^{2}(0, T ; H), \theta_{\varepsilon} \rightarrow \widehat{\theta}$ слабо в $L^{2}(0, T ; V)$,
    сильно в $L^{2}(0, T ; H)$;
    \[
        \int_{0}^{T} \int_{G_{b}} F\left(\theta_{\varepsilon}\right) d x d t \rightarrow \int_{0}^{T}
        \int_{G_{b}} F(\widehat{\theta}) dx dt \quad \text { и } \quad \int_{0}^{T} \int_{G_{b}}
        F\left(\theta_{\varepsilon}\right) dx dt \rightarrow 0, \text { как } \varepsilon \rightarrow+0
    \]

    Следовательно, $F(\widehat{\theta})=0$ и $\left.\widehat{\theta}\right|_{G_{b}} \leq \theta_{*}$.
    Результатов сходимости достаточно, чтобы перейти к пределу как $\varepsilon \rightarrow+0$
    в системе состояний (6) и доказать, что предельная пара
    $\{\widehat{\theta}, \widehat{u}\} \in$ $L^{2}(0, T ; V) \times U_{a d}$
    является приемлемым для проблемы(CP).
    Поскольку функционал $J$ является слабо полунепрерывным снизу, то есть
    \[
        j \leq J(\widehat{\theta}) \leq \liminf J\left(\theta_{\varepsilon}\right) \leq J(\theta)=j=\inf J
    \]
    Тогда пара $\{\widehat{\theta}, \widehat{u}\}$ это решение проблемы $(\mathrm{CP})$.
    \textbf{Теорема 3.} Пусть выполняются условия (i)-(iv),
    $\theta_{0} \leq \theta_{*}$ a.e. в $\Omega, \theta_{b} \leq \theta_{*}$ a.e. в
    $\Sigma$. If $\left\{\theta_{\varepsilon}, u_{\varepsilon}\right\}$  решения проблемы
    $\left(C P_{\varepsilon}\right)$ for $\varepsilon>0$, тогда существует последовательность вида
    $\varepsilon \rightarrow+0$ $u_{\varepsilon} \rightarrow \widehat{u}$ слабо в
    $L^{2}(0, T ; H), \quad \theta_{\varepsilon} \rightarrow \widehat{\theta}$ слабо в $L^{2}(0, T ; V)$,
    сильно в $L^{2}(0, T ; H)$, where $\{\widehat{\theta}, \widehat{u}\}$ есть решение проблемы (CP).
